---
title: "Introduction to data"
author: ""
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(openintro)
```

Some define statistics as the field that focuses on turning information into knowledge. The first step in that process is to summarize and describe the raw information -- the data. In this lab we explore flights, specifically a random sample of domestic flights that departed from the three major New York City airports in 2013. We will generate simple graphical and numerical summaries of data on these flights and explore delay times. Since this is a large data set, along the way you'll also learn the indispensable skills of data processing and subsetting.


## Getting started

### Load packages

In this lab, we will explore and visualize the data using the **tidyverse** suite of packages. The data can be found in the companion package for OpenIntro labs, **openintro**.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
```

### The data

The [Bureau of Transportation Statistics](http://www.rita.dot.gov/bts/about/) (BTS) is a statistical agency that is a part of the Research and Innovative Technology Administration (RITA). As its name implies, BTS collects and makes transportation data available, such as the flights data we will be working with in this lab.

First, we'll view the `nycflights` data frame. Type the following in your console to load the data:

```{r load-data}
data(nycflights)
```

The data set `nycflights` that shows up in your workspace is a *data matrix*, with each row representing an *observation* and each column representing a *variable*. R calls this data format a **data frame**, which is a term that will be used throughout the labs. For this data set, each *observation* is a single flight.

To view the names of the variables, type the command

```{r names}
names(nycflights)
```

This returns the names of the variables in this data frame. The **codebook** (description of the variables) can be accessed by pulling up the help file:

```{r help, eval=FALSE}
?nycflights
```

One of the variables refers to the carrier (i.e. airline) of the flight, which is coded according to the following system.

- `carrier`: Two letter carrier abbreviation.
    + `9E`:           Endeavor Air Inc.
    + `AA`:      American Airlines Inc.
    + `AS`:        Alaska Airlines Inc.
    + `B6`:             JetBlue Airways
    + `DL`:        Delta Air Lines Inc.
    + `EV`:    ExpressJet Airlines Inc.
    + `F9`:      Frontier Airlines Inc.
    + `FL`: AirTran Airways Corporation
    + `HA`:      Hawaiian Airlines Inc.
    + `MQ`:                   Envoy Air
    + `OO`:       SkyWest Airlines Inc.
    + `UA`:       United Air Lines Inc.
    + `US`:             US Airways Inc.
    + `VX`:              Virgin America
    + `WN`:      Southwest Airlines Co.
    + `YV`:          Mesa Airlines Inc.


Remember that you can use `glimpse` to take a quick peek at your data to understand its contents better.

```{r glimpse}
glimpse(nycflights)
```

The `nycflights` data frame is a massive trove of information. Let's think about some questions we might want to answer with these data:

- How delayed were flights that were headed to Los Angeles?
- How do departure delays vary by month?
- Which of the three major NYC airports has the best on time percentage for departing flights?


## Analysis


### Departure delays

Let's start by examing the distribution of departure delays of all flights with a histogram.

```{r hist-dep-delay}
ggplot(data = nycflights, aes(x = dep_delay)) +
  geom_histogram()
```

This function says to plot the `dep_delay` variable from the `nycflights` data frame on the x-axis. It also defines a `geom` (short for geometric object), which describes the type of plot you will produce. 

Histograms are generally a very good way to see the shape of a single distribution of numerical data, but that shape can change depending on how the data is split between the different bins. You can easily define the binwidth you want to use:

```{r hist-dep-delay-bins}
ggplot(data = nycflights, aes(x = dep_delay)) +
  geom_histogram(binwidth = 15)
ggplot(data = nycflights, aes(x = dep_delay)) +
  geom_histogram(binwidth = 150)
```

1.  Look carefully at these three histograms. How do they compare? Are features revealed in one that are obscured in another?

**Insert your answer here**
The changed binwidth allows for changing the appearance of the histograms. When comparing the first with the third, for example, it seems that the firt one shows more detail due to a smaller binwidth. The third histogram especially is too broad and does not allow for any clear examination. The second one has the smallest binwidth, and it can be nicely seen that there is a very strong bias towards one bin. The first histogram shows a bit broader binwidth to the second one, which in turn seenms to be the best to evaluate data distributiion. Generally, the binwidth choice depends on the spread the data has.

If you want to visualize only on delays of flights headed to Los Angeles, you need to first `filter` the data for flights with that destination (`dest == "LAX"`) and then make a histogram of the departure delays of only those flights.

```{r lax-flights-hist}
lax_flights <- nycflights %>%
  filter(dest == "LAX")
ggplot(data = lax_flights, aes(x = dep_delay)) +
  geom_histogram()
```

Let's decipher these two commands (OK, so it might look like four lines, but the first two physical lines of code are actually part of the same command. It's common to add a break to a new line after `%>%` to help readability).

- Command 1: Take the `nycflights` data frame, `filter` for flights headed to LAX, and save the result as a new data frame called `lax_flights`.
    + `==` means "if it's equal to".
    + `LAX` is in quotation marks since it is a character string.
- Command 2: Basically the same `ggplot` call from earlier for making a histogram, except that it uses the smaller data frame for flights headed to LAX instead of all flights.

<div id="boxedtext">
**Logical operators: ** Filtering for certain observations (e.g. flights from a particular airport) is often of interest in data frames where we might want to examine observations with certain characteristics separately from the rest of the data. To do so, you can use the `filter` function and a series of **logical operators**. The most commonly used logical operators for data analysis are as follows:

- `==` means "equal to"
- `!=` means "not equal to"
- `>` or `<` means "greater than" or "less than"
- `>=` or `<=` means "greater than or equal to" or "less than or equal to"
</div>

You can also obtain numerical summaries for these flights:

```{r lax-flights-summ}
lax_flights %>%
  summarise(mean_dd   = mean(dep_delay), 
            median_dd = median(dep_delay), 
            n         = n())
```

Note that in the `summarise` function you created a list of three different numerical summaries that you were interested in. The names of these elements are user defined, like `mean_dd`, `median_dd`, `n`, and you can customize these names as you like (just don't use spaces in your names). Calculating these summary statistics also requires that you know the function calls. Note that `n()` reports the sample size.

<div id="boxedtext">
**Summary statistics: ** Some useful function calls for summary statistics for a single numerical variable are as follows:

- `mean`
- `median`
- `sd`
- `var`
- `IQR`
- `min`
- `max`

Note that each of these functions takes a single vector as an argument and returns a single value. 
</div>

You can also filter based on multiple criteria. Suppose you are interested in flights headed to San Francisco (SFO) in February:

```{r sfo-feb-flights}
sfo_feb_flights <- nycflights %>%
  filter(dest == "SFO", month == 2)
```

Note that you can separate the conditions using commas if you want flights that are both headed to SFO **and** in February. If you are interested in either flights headed to SFO **or** in February, you can use the `|` instead of the comma.

2.  Create a new data frame that includes flights headed to SFO in February, 
    and save this data frame as `sfo_feb_flights`. How many flights 
    meet these criteria? 

```{r sfo-feb-flights2}
sfo_feb_flights <- nycflights %>%
  filter(dest == "SFO", month == 2)
```

There were 68 flights headed to SFO from NYC in Feb 2013.

3.  Describe the distribution of the **arrival** delays of these flights using a 
    histogram and appropriate summary statistics. **Hint:** The summary 
    statistics you use should depend on the shape of the distribution.
    
```{r arrival delays hist}
hist(sfo_feb_flights$arr_delay)
```
The histogram shows that the data is not really normally distributed, but also not extremely skewed in either direction; perhaps a bit to the right if we're strict. Given this distribution, investigating the mean is good practice, however, comparing the mean to the median never hurts and so I will compute several summary statistics and evaluate. Generally, it seems that the bulk of flights arriving at SFO is actually early.
This is unsurprising since many fligth times are actually overestimated to include common taxi times delays to runways.

```{r arrival delays summary stats}
summary(sfo_feb_flights$arr_delay)
```
Using the summary function, we can see that the mean is at -4.50 minutes and the median is at -11 minutes. This shows several things: firstly, the data is not very skewed as the mean and median are close together, and secondly, the average arrival delay is to arrive early. However, it also shows that there is sort of a physical limit on how early a flight can arrive early, whereas the delay can potentially be infinite. This leads to a big skewed data where we have longer delays than early arrvials.

Another useful technique is quickly calculating summary statistics for various groups in your data frame. For example, we can modify the above command using the `group_by` function to get the same summary stats for each origin airport:

```{r summary-custom-list-origin}
sfo_feb_flights %>%
  group_by(origin) %>%
  summarise(median_dd = median(dep_delay), iqr_dd = IQR(dep_delay), n_flights = n())
```

Here, we first grouped the data by `origin` and then calculated the summary statistics.

4.  Calculate the median and interquartile range for `arr_delay`s of flights in
    in the `sfo_feb_flights` data frame, grouped by carrier. Which carrier
    has the most variable arrival delays?

**Insert your answer here**
```{r sfo summaries}
sfo_feb_flights %>%
  group_by(carrier) %>%
  summarise(median_sfo = median(arr_delay),
            IQR_sfo = IQR(arr_delay))
```
Looking at the arrival delays grouped by carrier, we can see that American Airlines has the greatest delay, a median of 5 minutes, while Virgin America has the lowest, with a median of arriving 22.5 minutes early. Delta and United Airlines both have the highest interquartile range, of 22 minutes, suggesting that they have the greatest variability in arrival delays into SFO.


### Departure delays by month

Which month would you expect to have the highest average delay departing from an NYC airport?

Let's think about how you could answer this question:

- First, calculate monthly averages for departure delays. With the new language you are learning, you could
    + `group_by` months, then
    + `summarise` mean departure delays.
- Then, you could to `arrange` these average delays in `desc`ending order

```{r mean-dep-delay-months}
nycflights %>%
  group_by(month) %>%
  summarise(mean_dd = mean(dep_delay)) %>%
  arrange(desc(mean_dd))
```

5.  Suppose you really dislike departure delays and you want to schedule 
    your travel in a month that minimizes your potential departure delay leaving 
    NYC. One option is to choose the month with the lowest mean departure delay.
    Another option is to choose the month with the lowest median departure delay. 
    What are the pros and cons of these two choices?

**Insert your answer here**
The pro of using the median over the mean as a measure of centrality is that we can be sure the distribution of the month's delays won't affect our measure, making it a bit better to compare across months. However, in this case we actually care about the distribution, and how often such extreme delays occur, so the mean is a better measurement to understand that part better.

### On time departure rate for NYC airports

Suppose you will be flying out of NYC and want to know which of the three major NYC airports has the best on time departure rate of departing flights. Also supposed that for you, a flight that is delayed for less than 5 minutes is basically "on time."" You consider any flight delayed for 5 minutes or more to be "delayed".

In order to determine which airport has the best on time departure rate, you can 

- first classify each flight as "on time" or "delayed",
- then group flights by origin airport,
- then calculate on time departure rates for each origin airport,
- and finally arrange the airports in descending order for on time departure percentage.

Let's start with classifying each flight as "on time" or "delayed" by creating a new variable with the `mutate` function.

```{r dep-type}
nycflights <- nycflights %>%
  mutate(dep_type = ifelse(dep_delay < 5, "on time", "delayed"))
```

The first argument in the `mutate` function is the name of the new variable we want to create, in this case `dep_type`. Then if `dep_delay < 5`, we classify the flight as `"on time"` and `"delayed"` if not, i.e. if the flight is delayed for 5 or more minutes.

Note that we are also overwriting the `nycflights` data frame with the new version of this data frame that includes the new `dep_type` variable.

We can handle all of the remaining steps in one code chunk:

```{r ot-dep-rate}
nycflights %>%
  group_by(origin) %>%
  summarise(ot_dep_rate = sum(dep_type == "on time") / n()) %>%
  arrange(desc(ot_dep_rate))
```

6.  If you were selecting an airport simply based on on time departure percentage, which NYC airport would you choose to fly out of?

You can also visualize the distribution of on on time departure rate across the three airports using a segmented bar plot.

```{r viz-origin-dep-type}
ggplot(data = nycflights, aes(x = origin, fill = dep_type)) +
  geom_bar()
```

**Insert your answer here**
When looking at the percentages, it appears that LGA would be the best airport to fly out of with 73% on-time departures. 
* * *

## More Practice

7.  Mutate the data frame so that it includes a new variable that contains the 
    average speed, `avg_speed` traveled by the plane for each flight (in mph).
    **Hint:** Average speed can be calculated as distance divided by
    number of hours of travel, and note that `air_time` is given in minutes.
    
**Insert your answer here**

```{r avg speed}
nycflights = nycflights %>%
  mutate(avg_speed = distance / air_time * 60)
```

8.  Make a scatterplot of `avg_speed` vs. `distance`. Describe the relationship
    between average speed and distance.
    **Hint:** Use `geom_point()`.

**Insert your answer here**
```{r speed plot}
ggplot(data = nycflights, aes(x = avg_speed, y = distance)) +
  geom_point()
```

Looking at the scatterplot showing average speed versus distance, we can see several things. One, domestic flights have generally a maximum distance of just above 2500 miles (likely somewhere on the Northern Westcoast), except for Hawaii flights that are 5000 miles long. Additionally, there are physical limitations to speed, which is why we see the average speed to be very similiar for fliights above 500 miles. There is one outlier, with an average speed of 700 miles that is possibly a measurement error, or they took a Concorde to fly that, because the distance is also below 1,000 miles. 

9.  Replicate the following plot. **Hint:** The data frame plotted only
    contains flights from American Airlines, Delta Airlines, and United
    Airlines, and the points are `color`ed by `carrier`. Once you replicate
    the plot, determine (roughly) what the cutoff point is for departure
    delays where you can still expect to get to your destination on time.

```{r plot-to-replicate, echo=FALSE, fig.show="asis", fig.width=7, fig.height=4}
dl_aa_ua <- nycflights %>%
  filter(carrier == "AA" | carrier == "DL" | carrier == "UA")
ggplot(data = dl_aa_ua, aes(x = dep_delay, y = arr_delay, color = carrier)) +
  geom_point()
```

**Insert your answer here**
```{r plot, fig.show="asis", fig.width=7, fig.height=4}
ggplot(data = nycflights, aes(x = dep_delay, y = arr_delay, color = carrier)) +
  geom_point()
```
Looking at all arlines on the scatterplot, it appears that the cutoff point for a delayed flight to be still on time is not more than 50 to 60 minutes. Anything above that will almost certainly lead to a delayed arrival
